---
title: "2. Data Curation Survival_analysis"
author: "Valentina Quintero Santofimio"
format:
  html:
    include-in-header:
      - file: github-corner.html
editor: visual
---

# The UK Biobank (synthetic) dataset

Throughout this project, we will use a synthetic version of the UK Biobank data. The UK Biobank is a large cohort data set composed of over 500,000 volunteers recruited across United Kingdom between 2006-2010. For this project, we will use data from 252,877 participants who have high quality lung function data to select participants with airflow obstruction (defined as FEV1/FVC \< LLN)

## Essential Variables

| **Event (death) variables** | **Description**                                                                                                                                                           | **Coding**                                           | **Outcome variable** | **Description**                                                                                                                                              | **Coding**                                         |
|------------|------------|------------|------------|------------|------------|
| cofdmain                    | Main cause of death coded by International Classification of Disease v10 (ICD10) This variable will be used later in the course to stratify the analysis by type of death | 0 = No event (alive/censored) 1 = Event (dead)       | AO.fev1fvc           | Airflow obstruction derived as: forced expiratory volume at 1 s/ forced vital capacity below the lower limit of normal (from NHANES equations) (FEV1/FVCLLN) | 0 = No airflow obstruction 1 = Airflow obstruction |
| Date of death               | Date of participant's death obtained from the National Death Registry                                                                                                     | As character but must be changed into numeric format |                      |                                                                                                                                                              |                                                    |

## Additional variables used in the multivariable (adjusted model)

| Covariates name | Description                                                     | Coding                   |
|------------------|-------------------------------------|------------------|
| eid             | Unique participant Identification Number for anonymised data    | Interger                 |
| Age             | Age of participant at recruitment (when they entered the study) | Interger                 |
| Sex             | Sex of participant                                              | Male, Female             |
| bmi             | Body Max Index of participant                                   | Numeric                  |
| UKB_centre      | UK Biobank Centre at which the participant was recruited        | Character                |
| Ethnicity       | Genetically-derived ethnicity of the participant                | White vs Other           |
| Smoking_status  | Smoking Status of the participant at recruitment                | Never, current, previous |

# 2. Data Curation

Note: the `Surv()` function (next chapter) in the `{survival}` package accepts different formatting of the event date e.g. `TRUE/FALSE`, where `TRUE` is event and `FALSE` is censored; `1/0` where `1` is event and `0` is censored; or `2/1` where `2` is event and `1` is censored.

Please make sure that the event is properly formatted. **In this example we will be using 0/1**

```{r, cache=TRUE}
# Load Data set 

death <- read.table("data/synthetic_data_ReCoDe.txt", header=TRUE, sep="\t")

#Months need to be in number and not character
death$monthofbirth <- as.character(death$monthofbirth)
death$monthofbirth[death$monthofbirth == "January"] <- "01"
death$monthofbirth[death$monthofbirth == "February"] <- "02"
death$monthofbirth[death$monthofbirth == "March"] <- "03"
death$monthofbirth[death$monthofbirth == "April"] <- "04"
death$monthofbirth[death$monthofbirth == "May"] <- "05"
death$monthofbirth[death$monthofbirth == "June"] <- "06"
death$monthofbirth[death$monthofbirth == "July"] <- "07"
death$monthofbirth[death$monthofbirth == "August"] <- "08"
death$monthofbirth[death$monthofbirth == "September"] <- "09"
death$monthofbirth[death$monthofbirth == "October"] <- "10"
death$monthofbirth[death$monthofbirth == "November"] <- "11"
death$monthofbirth[death$monthofbirth == "December"] <- "12"

#Format of year XXXX- MonthXX- day XX (As a date)
death$dateofbirth <- paste0(death$yearofbirth,"-", death$monthofbirth, "-01")
death$dateofbirth <- as.Date(death$dateofbirth)
death$date_of_death <- as.Date(death$date_of_death)
death$now_cens <- as.Date("2023-01-01") #Change to "end of study" censoring date
```

Now that the dates are formatted, we need to calculate the difference between start and end dates in some units, in this example the time will be calculated in years.

As participants could enter the study at any point between 2006-2010, there is a delayed entry into the study. We therefore use age as the time-scale to account for this.

```{r, cache=TRUE}
death$timetoevent <- death$date_of_death - death$dateofbirth 
death$timetocens <- death$now_cens - death$dateofbirth 

#Select final time if not dead then censoring date
death$time <- ifelse(is.na(death$date_of_death), death$timetocens, death$timetoevent)

death$time_years <- death$time/365.25 #To have time in years

##Censoring of participants- do we know when they died? if so = 1 (Event occurs)
##Otherwise = 0 as we do not know when this could happen. 

death$allcause_death <- as.character(death$cofdmain)
death$allcause_death[is.na(death$allcause_death)] <- 0 #No event occured (ALIVE)
death$allcause_death[death$allcause_death > 0] <- 1 #Event occurs (Death)
```

## Overview of the data

```{r}
library(knitr)

head_data <- head(death[, c("Sex","Smoking_status", "allcause_death", "time_years")], 15)

# Create a table
kable(head_data, caption = "First 15 Observations of Survival Analysis Data")

```

## How many participants have experienced the event? (e.g. dead)

```{r}

library(dplyr)
library(tidyr)
library(knitr)

# Create a summary table
summary_table <- death %>%
  group_by(Sex, allcause_death) %>%
  summarise(count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = Sex, values_from = count, values_fill = list(count = 0)) %>%
  mutate(Total = Male + Female)

# Display the summary table
kable(summary_table, caption = "Number of participants that have died by sex")

```

## How many participants have airflow obstruction (outcome)?

```{r}
library(dplyr)
library(tidyr)
library(knitr)

summary_table <- death %>%
  group_by(Sex, AO.fev1fvc) %>%
  summarise(count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = Sex, values_from = count, values_fill = list(count = 0)) %>%
  mutate(Total = Male + Female)

# Display the summary table
kable(summary_table, caption = "Number of participants with airflow obstruction by sex")

```

```{r, cache=TRUE}
save(death, file = "result.RData")

```
